Data Migration under a Changing Schema

### 2023-07-12 2022Migration/articleMigrationFACS.tex

%\def\id#1{\text{\it #1\/}}
\newcommand{\xrightarrowdbl}[2][]{%
  \xrightarrow[#1]{#2}\mathrel{\mkern-14mu}\rightarrow
}
\newcommand{\id}[1]{\text{\it #1\/}}
\newcommand{\code}[1]{\text{\tt\small #1}}
\newcommand{\stmtText}[1]{``{\small\tt #1}''}
\newcommand{\dom}[1]{\id{dom}(#1)}
\newcommand{\cod}[1]{\id{cod}(#1)}
%\renewcommand{\int}[2]{\id{inter}(#1,#2)}
\newcommand{\relsIn}[1]{\id{relsIn}(#1)}    % maps a Term to a set of Relations
\newcommand{\popF}[1]{\id{pop}_{#1}}
\newcommand{\pop}[2]{\popF{#1}(#2)}
\newcommand{\maintain}{\mathbin{\id{maint}}}
\newcommand{\enf}{\mathbin{{\tt enforce}}}
\newcommand{\enforce}[2]{{\tt enforce}_{#1}(#2)}
\newcommand{\instance}{\mathbin{\id{inst}}}
\newcommand{\relname}[1]{\id{relname}(#1)}
\newcommand{\evt}[2]{\id{event}_{#1,#2}}
\newcommand{\src}[1]{\id{src}(#1)}
\newcommand{\tgt}[1]{\id{tgt}(#1)}
\newcommand{\sat}[2]{\id{sat}_{#1}(#2)}
\newcommand{\viol}[2]{{#1}(#2)}
%\newcommand{\violC}[1]{\id{viol}_{#1}}
\newcommand{\sign}[1]{\id{sign}(#1)}
\newcommand{\powerset}[1]{\cal{P}\{#1\}}
\newcommand{\la}{\langle}
\newcommand{\ra}{\rangle}
\newcommand{\full}{V}
\newcommand{\declare}[3]{\id{#1}_{\pair{#2}{#3}}}
\newcommand{\subst}[3]{#3_{[#1\rightarrow #2]}}
\newcommand{\fullt}[2]{V_{\pair{#1}{#2}}}
\newcommand{\iden}{I}
\newcommand{\ident}[1]{I_{\id{#1}}}
\newcommand{\expr}[3]{(#1)_{#2\times #3}}
\newcommand{\pair}[2]{\langle{#1},{#2}\rangle}
\newcommand{\maprel}[2]{{\tt maprel}_{#1}({#2})}
\newcommand{\Pair}[2]{#1\times#2}
\newcommand{\pairs}[1]{\id{pairs}(#1)}
\newcommand{\triple}[3]{\langle{#1},{#2},{#3}\rangle}
\newcommand{\quadruple}[4]{\langle{#1},{#2},{#3},{#4}\rangle}
\newcommand{\atom}[1]{{\tt\small #1}}
\newcommand{\atoms}{\mathcal{A}}
\newcommand{\Atoms}{\mathbb{A}}
%\newcommand{\events}{\mathcal{E}}
%\newcommand{\Events}{\mathbb{E}}
\newcommand{\concept}[1]{{\tt\small #1}}
\newcommand{\concepts}{\mathcal{C}}
\newcommand{\Concepts}{\mathbb{C}}
\newcommand{\decls}{\mathcal{D}}  %% names of relations
\newcommand{\rels}{\mathcal{R}}   %% all relations
\newcommand{\Rels}{\mathbb{R}}   %% all relations
\newcommand{\relations}{\mathcal{M}} % representing terms. M is a subset of R.
\newcommand{\triples}{\mathcal{T}}
\newcommand{\Triples}{\mathbb{T}}
\newcommand{\Triple}[3]{#1\times#2\times#3}
\newcommand{\vertices}{N}
\newcommand{\enforces}{\mathcal{E}}
\newcommand{\rules}{\mathcal{U}}
\newcommand{\Rules}{\mathbb{U}}
\newcommand{\specrules}{\mathcal{S}}
\newcommand{\roles}{\mathcal{O}}
\newcommand{\dataset}{\mathscr{D}}
\newcommand{\Dataset}{\mathbb{D}}
\newcommand{\schema}{\mathscr{Z}}
\newcommand{\functionality}{\mathscr{F}}
\newcommand{\select}[2]{\id{select}_{#1}\{{#2}\}}
\newcommand{\migrsys}{\mathscr{M}}
\newcommand{\infsys}{\mathscr{S}}
\newcommand{\tf}[1]{\mathscr{T}(#1)}
\newcommand{\ptf}[1]{\mathscr{T}'(#1)}
\newcommand{\ti}[1]{\mathscr{I}(#1)}
\newcommand{\tic}[1]{I_{\cal C}(#1)}
\newcommand{\relAdd}{\dagger}
\newcommand{\flip}[1]{{#1}^\smallsmile} %formerly:  {#1}^\backsim
\newcommand{\kleeneplus}[1]{{#1}^+}
\newcommand{\kleenestar}[1]{{#1}^*}
\newcommand{\cmpl}[1]{\overline{#1}}
\newcommand{\rel}{\times}
\newcommand{\compose}{;}
\newcommand{\subs}{\subseteq}%{\models}
\newcommand{\fun}{\rightarrow}
\newcommand{\isa}{\preceq}
%\newcommand{\isaClos}{\sqsubseteq}
\newcommand{\typetest}{?}
\newcommand{\meet}{\sqcap}
\newcommand{\join}{\sqcup}
\newcommand{\Meet}{\bigsqcap}
\newcommand{\Moin}{\bigsqcup} % because LaTeX has already defined command \Join.
\newcommand{\order}{\ominus}
\newcommand{\anything}{\top}
\newcommand{\nothing}{\bot}
\newcommand{\rewriteto}{\rightarrow}
\newcommand{\calc}{\implies}
\newcommand{\alland}{\bigwedge}
\newcommand{\mph}[3]{#1_{#2\times #3}}
\newcommand{\mphu}[1]{#1_{\univ\times\univ}}

@@

\keywords{Generative software \and Incremental software deployment \and Data migration \and Relation algebra \and Ampersand \and Schema change.}
\end{abstract}
%
%
%
\section{Introduction}
\label{sct:Introduction}
   The purpose of this work is to automate the data migrations that come with incrementally developing and deploying information systems.
   This paper proposes a theory for such data migrations in the context of generated information systems%
\footnote{In the sequel, the word ``system'' refers to the phrase ``information system''. This simplifies the language a little. }.

@@

   The contribution of this paper is to derive a migration schema from two artifacts: an existing system
   (which has its own schema and data set) and the specification of a desired system (which has a different schema and will ``import'' relevant parts of the existing data).   
   Our theory is based on the following assumptions and requirements:
\begin{itemize}
   \item {\em incremental deployment}\\The data migration is meant to deploy a software increment in production.
   \item {\em data pollution}\\The existing data set may be polluted, but it satisfies its schema.
   \item {\em human intervention}\\The data migration may require human interaction, which may take time.
   %\item {\em semantic continuity}\\the meaning of data must be preserved.
   \item {\em zero down-time}\\The business continues during the migration without interruption.
   \item {\em generative software development}\\A compiler exists to generate an information system from a given schema.
\end{itemize}
   This paper focuses on the correctness of the migration.
   Efficiency of the migration is beyond the scope.

\section{Analysis}
\label{sct:Analysis}
\subsection{Information Systems}
   The purpose of an information system is to store and disclose data in a way that is meaningful to its users.
   This section defines information systems, after which we define a migration procedure from one information system to another.

@@

\subsection{Data sets}
\label{sct:Data sets}
   A data set $\dataset$ describes a set of structured data, which is typically stored persistently in a database of some kind.
   The notation $\dataset_{\infsys}$ refers to the data set of a particular information system $\infsys$.
   The purpose of a data set is to describe the data of a system at one point in time. 
   Before defining data sets, let us first define the constituent notions of atom, concept, specialization, relation, and triple.

@@

%    Specialization, also known as {\em generalization} or {\em subtyping}, is a relation between concepts.
%    The statement $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
% \begin{equation}
%    \label{eqn:specialization}
%    A\isa B\ \Leftrightarrow\ \forall a: a\instance A\rightarrow a\instance B
% \end{equation}
%    Specialization is needed to allow statements such as: ``An employee is a person'' or ``A human is a mammal''.
%    In her script, a user can declare a specialization as a pair of concepts in the relation $\isa$.
%    A compiler can construct $\isa$ as the transitive, reflexive closure of all user-defined specializations.
%    It must make sure that $\isa$ is antisymmetric, so that $\isa$ is a partial order of concepts.
%    Specialization causes atoms to have multiple concepts of which they can be an instance.
%    As a result, if an atom is an instance of concept $A$ and $A\isa B$,
%    this atom has all properties that atoms of type $B$ have.

@@

\subsection{Schemas}
\label{sct:Schemas}
   Schemas serve to capture the semantics of an information system~\cite{Spivak2012}.
   We consider a schema to represent the static part of an information system, defining concepts, relations, and rules.
   A software engineer defines a schema on design time, so that semantic checks can be implemented at compile time.

   We describe a schema $\schema$ via a tuple $\triple{\concepts}{\rels}{\rules}$,
   in which $\concepts\subseteq \Concepts$ is a finite set of concepts,
   $\rels\subseteq \Rels$ is a finite set of relations,
   and $\rules\subseteq \left(\Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}\right)$ is a finite set of rules.

@@

   For a different kind of constraint, the system is told what to do in order to fix it.
   Such constraints are called enforce rules.
   We just need to consider one type of enforce rule in this work, its syntax is as follows:

\begin{align}
      \text{\tt ENFORCE}\ r\ \text{\tt >:}\ \id{term}\label{enforce ins}
\end{align}

   Adding an enforce rule to the script of an information system specifies a function $u : \Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}$ with $\viol{u}{\dataset} = \pop{\id{term} - r}{\dataset}$, and the information system ensures that the rule is satisfied by adding a triple $\triple{x}{r}{y}$ for each $\pair{x}{y} \in \viol{u}{\dataset}$, thus maintaining $\pop{r}{\dataset} \supseteq \pop{\id{term}}{\dataset}$ as an invariant.

   Formally, enforce rules are pairs with $u$ a function and $r$ a relation, written $r \mapsfrom u$ or equivalently $r \mapsfrom \lambda \dataset.~ u(\dataset)$.
   The set of enforce rules in a schema is $\enforces$.

   \begin{definition}[Schema]
   A schema is a tuple $\quadruple{\concepts}{\rels}{\rules}{\enforces}$ that satisfies:
\begin{align}
   \declare{n}{A}{B}\in\rels&~\Rightarrow~ A\in\concepts\ \wedge\ B\in\concepts
   \label{eqn:relationsIntroduceConcepts}\\
   \pair{u}{r}\in\mathcal\enforces&~\Rightarrow~ r \in \rels\label{eqn:enforceRulesRels}
\end{align}
   \end{definition}
   Requirement~\ref{eqn:relationsIntroduceConcepts} ensures that concepts mentioned in a relation are defined in the schema.
   The type system of Ampersand~\cite{vdWoude2011} ensures that every information system it generates satisfies this requirement.
   This is also known as static typing, which has well established advantages for the software engineering process~\cite{HanenbergKRTS14,Petersen2014}.

   %The notation $\schema_{\infsys}$ refers to the schema of information system $\infsys$.
   When clarity is needed, we write $\concepts_{\schema}$, $\rels_{\schema}$, $\rules_{\schema}$, and $\enforces_{\schema}$
   for $\concepts$, $\rels$, $\rules$, and $\enforces$ corresponding to $\schema = \quadruple{\concepts}{\rels}{\rules}{\enforces}$.

%   Note that these requirements do not depend on any particular data set,
%   so they can be checked at compile time.
%   This is also known as static typing,
%   which has well established advantages for the software engineering process~\cite{HanenbergKRTS14,Petersen2014}.

%    For every rule $u$ in the schema, we assume there is a function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$
%    that represents the set of violations%
% \footnote{The definition of ``violation'' is not needed in this paper. We only use the absence of violations to show satisfaction of a constraint.}
%    of rule $u$ in any data set.
%    We also assume a function $\id{sign} : \Rules\rightarrow\Pair{\Concepts}{\Concepts}$
%    that represents the signature of a rule, i.e. the type of atoms in violations:
% \begin{eqnarray}
%    \sign{u}=\pair{A}{B}\ \wedge\ \pair{a}{b}\in\viol{u}{\dataset}&\Rightarrow&a\instance A\wedge b\instance B
%    \label{eqn:wellTypedEdged violations}
% \end{eqnarray}
%    To determine whether rule $u$ satisfies data set $\dataset$,
%    we define $\sat{u}{\Dataset}$:
% \begin{eqnarray}
%    \sat{u}{\dataset}&\Leftrightarrow&\viol{u}{\dataset}=\emptyset
%    \label{eqn:sat}
% \end{eqnarray}
%    Violations are used to produce meaningful error messages and to trigger changes in the data set to restore invariance of that rule.

   % The maintainance relation between roles and rules is used to determine who is allowed to change the data set.
   % For every rule $u$ in the schema, there is a function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$
   % that represents the set of violations.%

   % In practice, there are different ways of enforcing rules: automatic enforcement and manual enforcement.

   % Automatic enforcement is specified by the developer with a special syntax:
   % the {\em enforcement rule}.
   % An enforcement rule specifies not only the rule,
   % but it implicitly defines event $e'$ to restore the invariance.
   % The system must have an engine that restores invariance of all enforcement rules,
   % so users will experience that these rules are always satisfied.

   % Manual enforcement means that the system (temporarily) allows violations,
   % to let a user ``invent'' a reaction $e'$ that restores invariance.
   % That is why the term $(\exists o\in\roles:\ o\maintain u)$ is added to requirement~\ref{eqn:satisfaction}.
   % It allows the compiler to generate code for users with role $o$,
   % so they can restore the invariant manually.
   % In the Ampersand language, a developer explicitly assigns a rule $u$ to a role $o$,
   % to keep control over the type of persons that are allowed to ``restore broken rules''.
   % As a consequence, a mechanism is needed to notify those persons of the work to be done.

   % If a rule $u$ is not assigned to any role and it is not an enforcement rule,
   % the system must prevent any event $e$ from having any effect on the database.
   % So, if $e$ would violate rule $u$, the system must reject $e$.
   % This implements blocking behaviour, and must be accompanied by an error message.

\subsection{Information Systems}
\label{sct:Information Systems}
%   An information system has a changing data set because of events that occur.
%   Whenever the system ``observes'' an event, it attempts to change its data set accordingly.
%   However, every rule in the schema represents a constraint,
%   which the data set must invariably satisfy.
%   That is why rules are also known as ``invariants''.
%
%   Events are taken from an infinite set, $\Events$.
%   An event is denoted as $\evt{d^-}{d^+}$, in which $d^-$ and $d^+$ are data sets.
%   Every event is a function $\Dataset\rightarrow\Dataset$, so $\evt{d^-}{d^+}(\dataset)$ is a data set.
%   Equation~\ref{eqn:event} defines the effect of an event on a data set.
%\begin{eqnarray}
%      \evt{d^-}{d^+}(\dataset)&=&\la(\triples_{\dataset}-\triples_{d^-})\cup\triples_{d^+},(\instance_{\dataset}-\instance_{d^-})\cup\instance_{d^+}\ra
%\label{eqn:event}
%\end{eqnarray}
%   In English, we would say that an event $\evt{d^-}{d^+}$ removes the triples in $d^-$ from $\dataset$ and then inserts the triples in $d^+$ into the result.
%   In the sequel, variables $e$, $e'$, $e''$, etc. represent events.
%   As soon as a system $\infsys$ observes an event $e$,
%   it applies $e$ to its data set $\dataset$.
%   Yet, if $e(\dataset)$ does not satisfy a rule $u$ from schema $\schema_\infsys$,
%   another event is needed, say $e'$, to ensure that $\sat{u}{e'(e(\dataset))}$.
%   The phrase ``to restore an invariant'' or ``to restore invariance of a rule''
%   means that the system will change $e(\dataset)$ by applying another event $e'$ to satisfy requirement~\ref{eqn:enforce}.
%   The notation $\enf_{u,e}$ represents this combination of events $e$ and $e'$.
%   So, $\enf_{u,e}$ must satisfy:
%\begin{eqnarray}
%   \sat{u}{\dataset}&\Rightarrow&\sat{u}{\enforce{u,e}{\dataset}}
%\label{eqn:enforce}
%\end{eqnarray}

   Let us now define the notion of an information system.
\begin{definition}[information system]
\label{def:information system}
\item An information system $\infsys$ is a tuple $\pair{\dataset}{\schema}$, in which
\begin{itemize}
   \item $\dataset=\pair{\triples}{\instance}$ is a data set (satisfies Equation~\ref{eqn:wellTypedEdge}), we write $\triples_\infsys = \triples$ and $\instance_\infsys = \instance$;
   \item $\schema=\triple{\concepts}{\rels}{\rules}$ is a schema (satisfies Equation~\ref{eqn:relationsIntroduceConcepts} and~\ref{eqn:enforceRulesRels}), we write $\concepts_\infsys = \concepts$, $\rels_\infsys = \rels$ and $\rules_\infsys=\rules$;
   \item all rules and enforce rules are satisfied:
   \begin{align}
   \forall u\in\rules&:\viol{u}{\dataset}=\emptyset\\
   \forall (r\mapsfrom u)\in\enforces&:\viol{u}{\dataset} - \pop{r}{\dataset}=\emptyset
   \label{eqn:satisfaction}
   \end{align}
   \item triples in the data set have their relation mentioned in the schema:
   \begin{eqnarray}
   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&\declare{n}{A}{B}\in\rels
   \label{eqn:define R}
   \end{eqnarray}
\end{itemize}
\end{definition}
   % A \define{role} is a name that identifies a group of users.
   % It serves as a placeholder for a person or a machine (i.e. an actor) that works with the data set (i.e. create, read, update, or delete triples).
   % The purpose of a role is to mention an individual user (human) or an automated actor (bot) without knowing who that user is.

   As the system responds to events, it changes its data set, while the schema remains the same.
   In this paper, it suffices to define an event as a pair of information systems for which the schema stays constant.
   Events are categorized by what part of the data set changes.

   \begin{definition}[Event]
   Let $R \subseteq \Rels$ and $I \subseteq \Rels$ be sets of relations, and let $\infsys' = \pair{\schema}{\dataset'}$ be an information system, and:
      \begin{align}
      \triples_{\dataset} - (\Triple{\Atoms}{(R \cup I)}{\Atoms}) &= \triples_{\dataset'} - (\Triple{\Atoms}{(R \cup I)}{\Atoms})
   \label{eqn:eventUnchanged}\\
      \triples_{\dataset} - (\Triple{\Atoms}{R}{\Atoms}) &\subseteq \triples_{\dataset'} - (\Triple{\Atoms}{R}{\Atoms})
   \label{eqn:eventInsert}
   \end{align}
   Then we say that $\pair{\pair{\schema}{\dataset}}{\infsys'}$ is an event with scope $R$ inserting in $I$.
   We use the notation $\pair{\schema}{\dataset} \xrightarrow[I]{R} \infsys'$ to indicate this.
   \end{definition}

   Equation~\ref{eqn:eventUnchanged} states that the triples of those for relations in $I$ or $R$ are the only ones to have changed, and Equation~\ref{eqn:eventInsert} states that of the ones that are only in $I$, no triples were removed.
   %If $I$ or $R$ is the empty set, we omit it from the arrow, so $\infsys \xrightarrow{R} \infsys'$ is a notation for $\infsys \xrightarrow[\emptyset]{R} \infsys'$.

@@

   We use inserting events to model the effect of enforce rules.
%   If $R$ is a set of relations that covers those that occur in the rule, then changes to an information system $\infsys$ can be described by the event $\infsys \xrightarrow[\{r\}]{R} \infsys'$.

\subsection{Example}
\label{sct:Example existing IS}
   Having defined an information system in mathematical terms, let us discuss a small example.
   It is written in the language Ampersand to make it more appealing to read.
   Let us first define a data set of just a handful of triples and three relations.
\begin{verbatim}
RELATION takes[Student*Course] =
[ ("Peter", "Management")
; ("Susan", "Business IT")
; ("John", "Business IT")
]
\end{verbatim}
   This declaration introduces a relation with the name \verb#takes#,
   source concept \verb#Student#, and
   target concept \verb#Course#.
   The informal meaning of this relation is that it states which students are taking which courses.

@@

   \subsection{Dealing with violations on invariants}
   Suppose the script for the desired system adds an invariant that is not present in the existing system:

@@

%   Let us proceed to specify a desired system, $\infsys'$, to illustrate a (toy) migration.

>>>>

%\def\id#1{\text{\it #1\/}}
\newcommand{\xrightarrowdbl}[2][]{%
  \xrightarrow[#1]{#2}\mathrel{\mkern-14mu}\rightarrow
}
\newcommand{\id}[1]{\text{\it #1\/}}
\newcommand{\code}[1]{\text{\tt\small #1}}
\newcommand{\stmtText}[1]{``{\small\tt #1}''}
\newcommand{\dom}[1]{\id{dom}(#1)}
\newcommand{\cod}[1]{\id{cod}(#1)}
%\renewcommand{\int}[2]{\id{inter}(#1,#2)}
\newcommand{\relsIn}[1]{\id{relsIn}(#1)}    % maps a Term to a set of Relations
\newcommand{\popF}[1]{\id{pop}_{#1}}
\newcommand{\pop}[2]{\popF{#1}(#2)}
\newcommand{\maintain}{\mathbin{\id{maint}}}
\newcommand{\enf}{\mathbin{{\tt enforce}}}
\newcommand{\enforce}[2]{{\tt enforce}_{#1}(#2)}
\newcommand{\instance}{\mathbin{\id{inst}}}
\newcommand{\relname}[1]{\id{relname}(#1)}
\newcommand{\evt}[2]{\id{event}_{#1,#2}}
\newcommand{\src}[1]{\id{src}(#1)}
\newcommand{\tgt}[1]{\id{tgt}(#1)}
\newcommand{\viol}[2]{{#1}(#2)}
%\newcommand{\violC}[1]{\id{viol}_{#1}}
\newcommand{\sign}[1]{\id{sign}(#1)}
\newcommand{\powerset}[1]{\cal{P}\{#1\}}
\newcommand{\la}{\langle}
\newcommand{\ra}{\rangle}
\newcommand{\full}{V}
\newcommand{\declare}[3]{\id{#1}_{\pair{#2}{#3}}}
\newcommand{\subst}[3]{#3_{[#1\rightarrow #2]}}
\newcommand{\fullt}[2]{V_{\pair{#1}{#2}}}
\newcommand{\iden}{I}
\newcommand{\ident}[1]{I_{\id{#1}}}
\newcommand{\expr}[3]{(#1)_{#2\times #3}}
\newcommand{\pair}[2]{\langle{#1},{#2}\rangle}
\newcommand{\maprel}[2]{{\tt maprel}_{#1}({#2})}
\newcommand{\Pair}[2]{#1\times#2}
\newcommand{\pairs}[1]{\id{pairs}(#1)}
\newcommand{\triple}[3]{\langle{#1},{#2},{#3}\rangle}
\newcommand{\quadruple}[4]{\langle{#1},{#2},{#3},{#4}\rangle}
\newcommand{\atom}[1]{{\tt\small #1}}
\newcommand{\atoms}{\mathcal{A}}
\newcommand{\Atoms}{\mathbb{A}}
%\newcommand{\events}{\mathcal{E}}
%\newcommand{\Events}{\mathbb{E}}
\newcommand{\concept}[1]{{\tt\small #1}}
\newcommand{\concepts}{\mathcal{C}}
\newcommand{\Concepts}{\mathbb{C}}
\newcommand{\decls}{\mathcal{D}}  %% names of relations
\newcommand{\rels}{\mathcal{R}}   %% all relations
\newcommand{\Rels}{\mathbb{R}}   %% all relations
\newcommand{\relations}{\mathcal{M}} % representing terms. M is a subset of R.
\newcommand{\triples}{\mathcal{T}}
\newcommand{\Triples}{\mathbb{T}}
\newcommand{\Triple}[3]{#1\times#2\times#3}
\newcommand{\vertices}{N}
\newcommand{\enforces}{\mathcal{E}}
\newcommand{\rules}{\mathcal{U}}
\newcommand{\Rules}{\mathbb{U}}
\newcommand{\specrules}{\mathcal{S}}
\newcommand{\roles}{\mathcal{O}}
\newcommand{\dataset}{\mathscr{D}}
\newcommand{\Dataset}{\mathbb{D}}
\newcommand{\schema}{\mathscr{Z}}
\newcommand{\functionality}{\mathscr{F}}
\newcommand{\select}[2]{\id{select}_{#1}\{{#2}\}}
\newcommand{\migrsys}{\mathscr{M}}
\newcommand{\infsys}{\mathscr{S}}
\newcommand{\tf}[1]{\mathscr{T}(#1)}
\newcommand{\ptf}[1]{\mathscr{T}'(#1)}
\newcommand{\ti}[1]{\mathscr{I}(#1)}
\newcommand{\tic}[1]{I_{\cal C}(#1)}
\newcommand{\relAdd}{\dagger}
\newcommand{\flip}[1]{{#1}^\smallsmile} %formerly:  {#1}^\backsim
\newcommand{\kleeneplus}[1]{{#1}^+}
\newcommand{\kleenestar}[1]{{#1}^*}
\newcommand{\cmpl}[1]{\overline{#1}}
\newcommand{\rel}{\times}
\newcommand{\compose}{;}
\newcommand{\subs}{\subseteq}%{\models}
\newcommand{\fun}{\rightarrow}
\newcommand{\isa}{\preceq}
%\newcommand{\isaClos}{\sqsubseteq}
\newcommand{\typetest}{?}
\newcommand{\meet}{\sqcap}
\newcommand{\join}{\sqcup}
\newcommand{\Meet}{\bigsqcap}
\newcommand{\Moin}{\bigsqcup} % because LaTeX has already defined command \Join.
\newcommand{\order}{\ominus}
\newcommand{\anything}{\top}
\newcommand{\nothing}{\bot}
\newcommand{\rewriteto}{\rightarrow}
\newcommand{\calc}{\implies}
\newcommand{\alland}{\bigwedge}
\newcommand{\mph}[3]{#1_{#2\times #3}}
\newcommand{\mphu}[1]{#1_{\univ\times\univ}}

@@

\keywords{Generative software \and Incremental software deployment \and Data migration \and Relation algebra \and Ampersand \and Schema change.}
\end{abstract}
%
%
%
\section{Introduction}
\label{sct:Introduction}
   The purpose of this work is to automate the data migrations that come with incrementally developing and deploying information systems.
   This paper proposes a theory for such data migrations in the context of generated information systems.
   In the sequel, the word ``system'' refers to the phrase ``information system'', to simplify the language a little.

@@

   The contribution of this paper is to derive a migration schema from two artifacts: an existing system
   (which has its own schema and data set) and the specification of a desired system (which has a different schema and will ``import'' relevant parts of the existing data).   
   Our theory is based on the following assumptions and requirements:
\begin{itemize}
   \item {\em incremental deployment}\\The data migration is meant to deploy a software increment in production.
   \item {\em data pollution}\\The existing data set may be polluted, but it satisfies its schema.
   \item {\em human intervention}\\The data migration may require human interaction, which may take time.
   %\item {\em semantic continuity}\\the meaning of data must be preserved.
   \item {\em zero down-time}\\The business continues during the migration without interruption.
   \item {\em generative software development}\\A compiler exists to generate a system from a given schema.
\end{itemize}
   This paper focuses on the correctness of the migration.
   Efficiency of the migration is beyond the scope.

\section{Analysis}
\label{sct:Analysis}
\subsection{Information Systems}
   The purpose of an information system is to store and disclose data in a way that is meaningful to its users.
   This section defines information systems, after which we define a migration procedure from one system to another.

@@

\subsection{Data sets}
\label{sct:Data sets}
   A data set $\dataset$ describes a set of structured data, which is typically stored persistently in a database of some kind.
   The notation $\dataset_{\infsys}$ refers to the data set of a particular system $\infsys$.
   The purpose of a data set is to describe the data of a system at one point in time. 
   Before defining data sets, let us first define the constituent notions of atom, concept, specialization, relation, and triple.

@@

@@

\subsection{Schemas}
\label{sct:Schemas}
   Schemas serve to capture the semantics of an information system~\cite{Spivak2012}.
   We consider a schema to represent the static part of a system, defining concepts, relations, and rules.
   A software engineer defines a schema on design time, so that semantic checks can be implemented at compile time.

   We describe a schema $\schema$ via a tuple $\triple{\concepts}{\rels}{\rules}$,
   in which $\concepts\subseteq \Concepts$ is a finite set of concepts,
   $\rels\subseteq \Rels$ is a finite set of relations,
   and $\rules\subseteq \powerset{\Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}}$ is a finite set of rules.

@@

   For a different kind of constraint, the system is told what to do in order to fix it.
   Such constraints are called enforcement rules.
   We just use one type of enforcement rule\footnote{Ampersand has several types of enforcement rules.} in this work.
   Its syntax is as follows:
\begin{align}
      \text{\tt ENFORCE}\ r\ \text{\tt >:}\ \id{term}\label{enforce ins}
\end{align}

   Adding an enforcement rule to the script of a system specifies a function $u : \Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}$
   with $\viol{u}{\dataset} = \pop{\id{term} - r}{\dataset}$.
   The system ensures that the rule is satisfied by adding a triple $\triple{x}{r}{y}$ for each $\pair{x}{y} \in \viol{u}{\dataset}$.
   Thus the system maintains $\pop{r}{\dataset} \supseteq \pop{\id{term}}{\dataset}$ as an invariant.

   Formally, enforcement rules are pairs with $u$ a function and $r$ a relation, written $r \mapsfrom u$ or equivalently $r \mapsfrom \lambda \dataset.~ u(\dataset)$.
   The set of enforcement rules in a schema is $\enforces$.

   \begin{definition}[Schema]
   A schema is a tuple $\quadruple{\concepts}{\rels}{\rules}{\enforces}$ that satisfies:
\begin{align}
   \declare{n}{A}{B}\in\rels&~\Rightarrow~ A\in\concepts\ \wedge\ B\in\concepts
   \label{eqn:relationsIntroduceConcepts}\\
   \pair{u}{r}\in\mathcal\enforces&~\Rightarrow~ r \in \rels\label{eqn:enforceRulesRels}
\end{align}
   \end{definition}
   Requirement~\ref{eqn:relationsIntroduceConcepts} ensures that concepts mentioned in a relation are defined in the schema.
   The type system of Ampersand~\cite{vdWoude2011} ensures that every system it generates satisfies this requirement.
   This is also known as static typing, which has well established advantages for the software engineering process~\cite{HanenbergKRTS14,Petersen2014}.

   %The notation $\schema_{\infsys}$ refers to the schema of system $\infsys$.
   When clarity is needed, we write $\concepts_{\schema}$, $\rels_{\schema}$, $\rules_{\schema}$, and $\enforces_{\schema}$
   for $\concepts$, $\rels$, $\rules$, and $\enforces$ corresponding to $\schema = \quadruple{\concepts}{\rels}{\rules}{\enforces}$.

\subsection{Information Systems}
\label{sct:Information Systems}
   Let us now define the notion of an information system.
\begin{definition}[information system]
\label{def:information system}
\item An information system $\infsys$ is a tuple $\pair{\dataset}{\schema}$, in which
\begin{itemize}
   \item $\dataset=\pair{\triples}{\instance}$ is a data set (satisfies equation~\ref{eqn:wellTypedEdge}), we write $\triples_\infsys = \triples$ and $\instance_\infsys = \instance$;
   \item $\schema=\quadruple{\concepts}{\rels}{\rules}{\enforces}$ is a schema (satisfies equation~\ref{eqn:relationsIntroduceConcepts} and~\ref{eqn:enforceRulesRels}), we write $\concepts_\infsys = \concepts$, $\rels_\infsys = \rels$, $\rules_\infsys=\rules$, and $\enforces_\infsys=\enforces$;
   \item all rules and enforcement rules are satisfied:
   \begin{align}
   \forall u\in\rules&.~\viol{u}{\dataset}=\emptyset\\
   \forall (r\mapsfrom u)\in\enforces&.~\viol{u}{\dataset} - \pop{r}{\dataset}=\emptyset
   \label{eqn:satisfaction}
   \end{align}
   \item triples in the data set have their relation mentioned in the schema:
   \begin{eqnarray}
   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&\declare{n}{A}{B}\in\rels
   \label{eqn:define R}
   \end{eqnarray}
\end{itemize}
\end{definition}
   % A \define{role} is a name that identifies a group of users.
   % It serves as a placeholder for a person or a machine (i.e. an actor) that works with the data set (i.e. create, read, update, or delete triples).
   % The purpose of a role is to mention an individual user (human) or an automated actor (bot) without knowing who that user is.

   As the system responds to events, it changes its data set, while the schema remains the same.
   In this paper, it suffices to define an event as a pair of systems for which the schema stays constant.
   Events are categorized by what part of the data set changes.

\begin{definition}[Event]
   Let $R \subseteq \Rels$ and $I \subseteq \Rels$ be sets of relations, and let $\infsys=\pair{\schema}{\dataset}$ and  $\infsys'=\pair{\schema}{\dataset'}$ be information systems, and:
      \begin{align}
      \triples_{\dataset} - (\Triple{\Atoms}{(R \cup I)}{\Atoms}) &= \triples_{\dataset'} - (\Triple{\Atoms}{(R \cup I)}{\Atoms})
   \label{eqn:eventUnchanged}\\
      \triples_{\dataset} - (\Triple{\Atoms}{R}{\Atoms}) &\subseteq \triples_{\dataset'} - (\Triple{\Atoms}{R}{\Atoms})
   \label{eqn:eventInsert}
   \end{align}
%   Then we say that $\pair{\pair{\schema}{\dataset}}{\infsys'}$ is an event with scope $R$ inserting in $I$.
   Then we say that $\infsys\xrightarrow[I]{R} \infsys'$ is an event
   that brings the system $\infsys$ to $\infsys'$ inserting pairs in relations from $I$ and updating (i.e. deleting and inserting) pairs in $R$.
%   We use the notation $\pair{\schema}{\dataset} \xrightarrow[I]{R} \infsys'$ to indicate this.
\end{definition}

   equation~\ref{eqn:eventUnchanged} states that the triples of those for relations in $I$ or $R$ are the only ones to have changed, and equation~\ref{eqn:eventInsert} states that of the ones that are only in $I$, no triples were removed.
   %If $I$ or $R$ is the empty set, we omit it from the arrow, so $\infsys \xrightarrow{R} \infsys'$ is a notation for $\infsys \xrightarrow[\emptyset]{R} \infsys'$.

@@

   We use inserting events to model the effect of enforcement rules.
%   If $R$ is a set of relations that covers those that occur in the rule, then changes to an system $\infsys$ can be described by the event $\infsys \xrightarrow[\{r\}]{R} \infsys'$.

\subsection{Example}
\label{sct:Example existing IS}
   Having defined information systems in mathematical terms, let us discuss a small example.
   It is written in the language Ampersand to make it more appealing to read.
   Let us first define a data set of just a handful of triples and three relations.
\begin{verbatim}
RELATION takes[Student*Course] =
[ ("Peter", "Management")
; ("Susan", "Business IT")
; ("John", "Business IT")
]
\end{verbatim}
   This declaration introduces a relation with the name \verb#takes#,
   source concept \verb#Student#, and
   target concept \verb#Course#.
   The informal meaning of this relation is that it states which students are taking which courses.

@@

\subsection{Dealing with violations on invariants}\label{sec:violations on invariants}
   Suppose the script for the desired system adds an invariant that is not present in the existing system:

@@
